---
title: "MY457: Problem Set Template"
date:  "`r format(Sys.time(), '%a/%d/%b')`"
author: "[Your Name Here]"
always_allow_html: true
output: 
  bookdown::pdf_document2:
    toc: false
---

```{r setup, include = FALSE}
# this chunk contains code that sets global options for the entire .Rmd. 
# we use include=FALSE to suppress it from the top of the document, but it will still appear in the appendix. 
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, linewidth=60)

# you can include your libraries here:
library(tidyverse)

# and any other options in R:
options(scipen=999)

```

# Concepts

We going to talk about a simple study of the effect of treatment $$
D_i \in \{0,1\} \text{ on } Y_i \text{ for all } i \in \{1,2,3,\dots,N\}
$$

## 1.1

$Y_{i1}$ means the potential outcome of the i unit observation in our treatment group. This situation of $Y_{i}$ having the same quantity as $Y_{i1}$ is only possible when all of our units of i are treated.

## 1.2

The difference between $E[Y_{i0} \mid D_i = 1]$ and $E[Y_{i0} \mid D_i = 0]$ is that the first one cannot be observed, while the second one we can observe.

I would expect these quantities to be equal when we randomly select the units to be treated. This is because the treatment is randomly assigned, so the potential outcomes are also randomly assigned.

We would expect these to be unequal when the treatment is not randomly assigned. This is because the potential outcomes are not randomly assigned, so the treatment is not randomly assigned.

## 1.3

Random assignment into our treatment (D=1) and control (D=0) groups ensures that the characteristics of the population of interest are also randomly distributed. This helps us identify the average treatment effect (ATE) because we can compare the average outcomes of the treatment and control groups and we dont have to worry about the confounding variables.

\clearpage

# Simulations

## 2.1





# Code appendix

```{r ref.label = knitr::all_labels(), echo=TRUE, eval=FALSE}
# this chunk generates the complete code appendix. 
# eval=FALSE tells R not to re-run (``evaluate'') the code here. 
```
