---
title: "MY457: Problem Set Template"
date:  "`r format(Sys.time(), '%a/%d/%b')`"
author: "[Your Name Here]"
always_allow_html: true
output: 
  bookdown::pdf_document2:
    toc: false
---

```{r setup, include = FALSE}
# this chunk contains code that sets global options for the entire .Rmd. 
# we use include=FALSE to suppress it from the top of the document, but it will still appear in the appendix. 
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, linewidth=60)

# you can include your libraries here:
library(tidyverse)
install.packages("report")
library(report)

# and any other options in R:
options(scipen=999)

```

# 1 Concepts

We going to talk about a simple study of the effect of treatment $$
D_i \in \{0,1\} \text{ on } Y_i \text{ for all } i \in \{1,2,3,\dots,N\}$$

## 1.1

$Y_{i1}$ means the potential outcome of the i unit observation in our treatment group. This situation of $Y_{i}$ having the same quantity as $Y_{i1}$ is only possible when all of our units of i are treated.

## 1.2

The difference between $E[Y_{i0} \mid D_i = 1]$ and $E[Y_{i0} \mid D_i = 0]$ is that the first one cannot be observed, while the second one we can observe.

I would expect these quantities to be equal when we randomly select the units to be treated. This is because the treatment is randomly assigned, so the potential outcomes are also randomly assigned.

We would expect these to be unequal when the treatment is not randomly assigned. This is because the potential outcomes are not randomly assigned, so the treatment is not randomly assigned.

## 1.3

By using a random assignment for our treatment (D=1) and control (D=0) groups helps us identify the average treatment effect (ATE) because we can compare the group level averages of potential outcomes of treatment and control and we dont have to worry about selection bias.

\clearpage

# 2 Simulations

## 2.1

```{r generate-data}
set.seed(123)
n <- 500
tau <- 5000

data <- data.frame(
Age = rnorm(n, mean = 42, sd = 10),
Education = sample(1:4, n, replace = TRUE),
Y0 = rnorm(n, mean = 50000, sd = 10000)
)
data <- data %>% mutate(
Y1 = Y0 + tau,
D = sample(c(0, 1), n, replace = TRUE),
Y = ifelse(D== 1, Y1, Y0)
)

data
```

The code above generates a dataset with 500 observations (n), a treatment effect of 5000 (tau), and four variables: Age with mean of 42 years of age with a standard deviation of 10 years, Education with 4 different categories represented by numbers 1 to 4, the binary variable D which is the treatment assignment and Y outcome variable (Y). Because we simulated the data we can also simulate both potential outcomes for each observation which are the Y1 and Y0 variables. We set the seed to 123 to ensure reproducibility.

## 2.2

```{r explore-data, warning=FALSE, fig.width=8, fig.height=6}}
# Perform t-test for Age
t_test_age <- t.test(Age ~ D, data = data)

# Perform t-test for Education
t_test_edu <- t.test(Education ~ D, data = data)

# Print results using report package
report::report(t_test_age)
report::report(t_test_edu)

```

Both Age and Education are balanced across treatment (D=1) and control (D=0) groups, as there are no statistically significant differences in means. This supports the randomization assumption that the two groups are comparable.

## 2.3

First lets plot the distribution of the potential outcomes $Y_0$ and $Y_1$.

```{r visualise-potential-outcomes, fig.width=8, fig.height=6}
data %>% 
  ggplot(aes(x = Y0)) +
  geom_density(aes(fill = "Y0"), alpha = 0.5) +
  geom_density(aes(x = Y1, fill = "Y1"), alpha = 0.5) +
  theme_minimal() +
  labs(title = "Potential Outcomes", x = "Y", y = "Density")
```

To figure out the difference in means between the potential outcomes $Y_0$ and $Y_1$ we can use the following code:

```{r difference-in-means}
# Calculate the difference in means between Y1 and Y0
diffINmeans <- mean(data$Y[data$D==1]) - mean(data$Y[data$D==0])
diffINmeans
```

Given that the treatment effect (tau) was set to 5000, but the estimated effect is 4181.018, the estimate is  lower than the true effect. The discrepancy may be due to randomization variability or sampling variability.
```{r}
report::report(t.test(Y ~ D, data = data))
```

## 2.4

```{r visualise-the-distribution of-Y-by-treatment-status, fig.width=8, fig.height=6}
# Plot the distribution of Y conditional on D
ggplot(data, aes(x = Y, fill = as.factor(D))) +
  geom_density(alpha = 0.5) +
  labs(title = "Distribution of Y by Treatment Status",
       x = "Outcome (Y)",
       y = "Density",
       fill = "Treatment (D)") +
  theme_minimal()
```

```{r}
diffInMeans <- mean(data$Y[data$D == 1]) - mean(data$Y[data$D == 0])
print(diffInMeans)

lm_model <- lm(Y ~ D, data = data)
summary(lm_model)

```

Both implementation gave the same 4181.018 results.  

## 2.5

```{r}
# Set parameters
n_simulations <- 1000
tau <- 5000
n <- 500

# Initialize vector to store deviations
ate_differences <- numeric(n_simulations)

# Run 1,000 simulations
for (i in 1:n_simulations) {
  # Generate new random dataset
  data <- data.frame(
    Age = rnorm(n, mean = 42, sd = 10),
    Education = sample(1:4, n, replace = TRUE),
    Y0 = rnorm(n, mean = 50000, sd = 10000)
  )
  
  # Apply intervention effects
  data <- data %>%
    mutate(
      Y1 = Y0 + tau,
      D = sample(c(0, 1), n, replace = TRUE),  # Randomly assign treatment
      Y = ifelse(D == 1, Y1, Y0)               # Observed outcome
    )
  
  # Estimate ATE (difference-in-means)
  estimated_ate <- mean(data$Y[data$D == 1]) - mean(data$Y[data$D == 0])
  
  # Store the difference from true ATE
  ate_differences[i] <- estimated_ate - tau
}

# Compute the mean of the differences
mean_diff <- mean(ate_differences)

# Plot histogram of ATE differences
ggplot(data.frame(ate_differences), aes(x = ate_differences)) +
  geom_histogram(color = "black", fill = "blue", bins = 30, alpha = 0.7) +
  geom_vline(aes(xintercept = mean_diff), color = "red", linetype = "dashed") +
  labs(
    title = "Distribution of Estimated ATE Differences",
    x = "Estimated ATE - True ATE",
    y = "Frequency"
  ) +
  theme_minimal()
```

After simulating 1000 data generations, the estimated ATE averageis really close to 0.

# 3. Replication

## 3.1

```{r}
#install.packages("readstata13")
library(readstata13)

# Load the dta data
electwomen_data <- read.dta13("how_to_elect_more_women.dta")

# Create sd_onefem2014 dummy variable from prop_sd_fem2014
electwomen_data <- electwomen_data %>%
  mutate(sd_onefem2014 = ifelse(prop_sd_fem2014 > 0, 1, 0))
```

## 3.2

```{r}
# Ensure there are no missing values
data <- electwomen_data %>% mutate(prop_sd_fem2014 = ifelse(is.na(prop_sd_fem2014), 0, prop_sd_fem2014))

# Group by 'condition' and calculate proportions correctly
condition_summary <- electwomen_data %>%
  group_by(condition) %>%
  summarise(
    precinct_count = n(),  # Number of precincts in each condition group
    sum_fem2014 = sum(prop_sd_fem2014, na.rm = TRUE),  # Sum of elected women proportion
    proportion_fem2014 = sum_fem2014 / sum(data$prop_sd_fem2014, na.rm = TRUE)  # Proportion per condition
  )

# Print results
print(condition_summary)
```

## 3.3

```{r}
#test balance in pre-treatment variables of "agecat" and "race" betwen different conditions
# Define pre-treatment variables
pre_treatment_vars <- c("agecat", "race")

# Define unique conditions (excluding NA)
conditions <- c("Control", "Demand", "Supply", "Supply+Demand")

# Function to perform pairwise t-tests
perform_t_test <- function(var, group1, group2) {
  test_result <- electwomen_data %>%
    filter(condition %in% c(group1, group2)) %>%
    t.test(reformulate("condition", var), data = .)
  
  cat("\nT-test for", var, "between", group1, "and", group2, "\n")
  print(test_result)
}

# Run t-tests for each variable between control and each treatment group
for (var in pre_treatment_vars) {
  for (group in conditions[conditions != "Control"]) {
    perform_t_test(var, "Control", group)
  }
}
```

No t test is significant, so we can say that the randomization was successful.

## 3.4

```{r}
library(dplyr)

# Function to run linear regression for ATE estimation
estimate_ate <- function(treatment_group) {
  # Subset to only include Control and the current treatment group
  subset_data <- electwomen_data %>%
    filter(condition %in% c("Control", treatment_group)) %>%
    mutate(treatment = ifelse(condition == treatment_group, 1, 0))  # Convert to binary

  # Run linear regression
  model <- lm(sd_onefem2014 ~ treatment, data = subset_data)
  
  # Print results
  cat("\nATE Estimate for", treatment_group, "vs. Control:\n")
  print(summary(model))
}

# Run regressions for each treatment group
treatment_groups <- c("Demand", "Supply", "Supply+Demand")

for (treatment in treatment_groups) {
  estimate_ate(treatment)
}

```

## 3.5



# Code appendix

```{r ref.label = knitr::all_labels(), echo=TRUE, eval=FALSE}
# this chunk generates the complete code appendix. 
# eval=FALSE tells R not to re-run (``evaluate'') the code here. 
```
